---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Loaded the activity data from the project directory into data-frame called mydata  
Created subset mydata1 with no missing values using the following code  

```{r echo=TRUE}
mydata <- read.csv("activity.csv", header=TRUE, sep=",")
mydata1 <- mydata[complete.cases(mydata),]
summary(mydata1)
```

## What is mean total number of steps taken per day?
Aggregate the steps per day and plot a histogram of steps taken per day   

```{r  Rough_Steps_Perday_Hist}
# Derive a vector mydata2 with sum of steps for each date
mydata2 <- tapply(mydata1$steps, mydata1$date, sum)

hist(mydata2, xlab= "Steps per day", main="Daily Steps Histogram")
```

Mean steps per day is: `r mean(mydata2, trim=0, na.rm=TRUE)`  
Median steps per day is: `r median(mydata2, na.rm=TRUE)`  

## What is the average daily activity pattern?
Derive Mean steps per 5 minute interval and plot as time series 

```{r Mean_steps_per_interval_ts}
mydata3 <- tapply(mydata1$steps, mydata1$interval, mean)


plot(mydata3, type="l", main=" Average Steps per 5min Interval", ylab="Steps", xlab="Interval")
# m1 <- max(mydata3)
# m1
m2 <- which.max(mydata3)
hh <- as.integer((m2-1)/12)
mm <- ((m2-1) - hh*12)*5
```

It can be seen that the maximum steps per interval is 'r max(mydata3)`.
This occurs in the interval commencing at `r hh` hour, `r mm` minutes. 


## Imputing missing values

The missing values of steps for any interval will be replaced by the average value of steps for that interval on other days.

The following code adds variables hh (hours), mint (minutes) and int (interval number) to the initial mydata data-frame. It then adds the mst variable (mean steps per interval). Then stepsx variable is added which is mst when there is a missing value for steps. 
```{r}
# Now deal with missing values
misr <- complete.cases(mydata)

# Add mean steps per interval as an additional column mst of the mydata dataframe
mydata$hh <-  as.integer(mydata$interval/100)
mydata$mint <-  as.integer((mydata$interval - mydata$hh*100)/5)
mydata$int <- (mydata$hh*12) + (mydata$mint + 1) 
# Look-up in mydata3 vector
mydata$mst <- mydata3[mydata$int]
mydata$stepsx <- ifelse(misr, mydata$steps, mydata$mst)
```

The following histogram shows the distribution of steps per day after imputation
```{r Imputed_Steps_Perday_Hist}
mydx <- tapply(mydata$stepsx, mydata$date, sum)
hist(mydx, xlab= "Steps per day", main="Steps Histogram after Imputation")

```

Mean steps per day using imputation is: `r mean(mydx, trim=0, na.rm=TRUE)`  
Median steps using imputation per day is: `r median(mydx, na.rm=TRUE)`  

So the impact of imputation on the mean is negligible

## Are there differences in activity patterns between weekdays and weekends?

To analyze the differences between weekend and weekday activity patterns, we add a character day variable and a logical weekend variable to the mydata dataframe 

```{r}
mydata$day <- weekdays(as.Date(mydata$date))
mydata$weekend <- ifelse((mydata$day=="Sunday") | (mydata$day == "Saturday"), TRUE, FALSE)
summary(mydata)
```

This supports display of activity time series for weekdays and weekends separately

```{r panel_weekend_weekday_activty_ts}
par(mfrow=c(2,1))

mydworkd <- mydata[!mydata$weekend, ]
summary(mydworkd)

```

